---
title: "data_eu dataset creation"
author: "Anna Anesiadou-Hansen"
date: "April 27, 2016"
output: html_document
---

Using different data sets from `http://www.gapminder.org/data/` I have build a data set called `data_eu`. Following files have been choosen:

* "Indicator_BMI female ASM.csv"
* "Indicator_BMI male ASM.csv"
* "Indicator_SBP female ASM.csv"
* "Indicator_SBP male ASM.csv"
* "Indicator_TC female ASM.csv"
* "Indicator_TC male ASM.csv"
* "indicator sugar_consumption.csv"
* "indicator food_consumption.csv"
* "indicator gapminder gdp_per_capita_ppp.csv"
* "indicator life_expectancy_at_birth.csv"

```{r echo = FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.
library(dplyr)
library(psych)
library(xlsx)
library(gridExtra)
library(tidyr)
library(memisc)

# Read BMI related .csv files, Reshape and Merge dataframes
BMI_female = read.csv("Indicator_BMI female ASM.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(BMI_female)[1] <- "Country"  # Rename the first column
BMI_female<- gather(BMI_female, Year, female, 2:ncol(BMI_female))  # Reshape dataframe
BMI_female$Year <- factor(BMI_female$Year, ordered = T) # Ordered column Year 

BMI_male = read.csv("Indicator_BMI male ASM.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(BMI_male)[1] <- "Country"  # Rename the first column
BMI_male<- gather(BMI_male, Year, male, 2:ncol(BMI_male))  # Reshape dataframe
BMI_male$Year <- factor(BMI_male$Year, ordered = T) # Ordered column Year

# Merge data sets BMI_female & BMI_male
BMI <- merge(BMI_female, BMI_male, by = c("Country", "Year")) %>% na.omit()
BMI<- gather(BMI, "Gender", "BMI_Index", 3:4)

#Underweight: BMI is less than 18.5
#Normal weight: BMI is 18.5 to 24.9
#Overweight: BMI is 25 to 29.9
#Obese: BMI is 30 or more

# Read Bloodpressure related .csv files, Reshape and Merge dataframes
Bloodpressure_female = read.csv("Indicator_SBP female ASM.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Bloodpressure_female)[1] <- "Country" # Rename the first column
Bloodpressure_female <- gather(Bloodpressure_female, Year, female, 2:ncol(Bloodpressure_female)) # Reshape dataframe
Bloodpressure_female$Year <- factor(Bloodpressure_female$Year , ordered = T) # Ordered column Year


Bloodpressure_male = read.csv("Indicator_SBP male ASM.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Bloodpressure_male)[1] <- "Country" # Rename the first column
Bloodpressure_male <- gather(Bloodpressure_male, Year, male, 2:ncol(Bloodpressure_male)) # Reshape dataframe
Bloodpressure_male$Year <- factor(Bloodpressure_male$Year , ordered = T) # Ordered column Year

Bloodpressure <- merge(Bloodpressure_female, Bloodpressure_male, by = c("Country","Year"))  %>% na.omit() # Merge data sets Bloodpressure_female & Bloodpressure_male
Bloodpressure <- gather(Bloodpressure, "Gender", "Bloodpressure", 3:4)

#Category 	                   systolic, mm Hg 	    diastolic, mm Hg
#Hypotension                        < 90                  < 60
#Desired                            90–119                60–79
#Prehypertension                    120–139               80–89
#Stage 1 hypertension               140–159               90–99
#Stage 2 hypertension               160–179               100–109
#Hypertensive urgency               ≥ 180                 ≥ 110
#Isolated systolic hypertension     ≥ 160                 < 90

# Read Cholesterol related .csv files, Reshape and Merge dataframes
Cholesterol_female <- read.csv("Indicator_TC female ASM.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Cholesterol_female)[1] <- "Country" # Rename the first column
Cholesterol_female <- gather(Cholesterol_female , Year, female, 2:ncol(Cholesterol_female)) # Reshape dataframe
Cholesterol_female$Year <- factor(Cholesterol_female$Year, ordered = T) # Ordered column Year

Cholesterol_male <- read.csv("Indicator_TC male ASM.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Cholesterol_male)[1] <- "Country" # Rename the first column
Cholesterol_male <- gather(Cholesterol_male , Year, male, 2:ncol(Cholesterol_male)) # Reshape dataframe
Cholesterol_male$Year <- factor(Cholesterol_male$Year, ordered = T) # Ordered column Year

Cholesterol<- merge(Cholesterol_female, Cholesterol_male, by = c("Country", "Year")) %>% na.omit()# Merge data sets Cholesterol_female & Cholesterol_male
Cholesterol<- gather(Cholesterol, "Gender", "Cholesterol", 3:4) # Reshape dataframe

# Level mg/dL 	    Level mmol/L 	     Interpretation
#   < 200 	            < 5.2 	          Desirable level corresponding to lower risk for heart disease
#   200–240 	          5.2–6.2 	        Borderline high risk
#   > 240 	              > 6.2 	        High risk



#Merge data sets BMI & Bloodpressure
BMI_Bloodpressure <- merge(BMI, Bloodpressure, by = c("Country","Year", "Gender")) %>% na.omit() 
BMI_Bloodpressure_Cholesterol <- merge(BMI_Bloodpressure, Cholesterol, by = c("Country", "Year", "Gender"))


# Create data set data
data <- BMI_Bloodpressure_Cholesterol 

# Read Sugar related .csv file, Reshape and Merge dataframe
Sugar <- read.csv("indicator sugar_consumption.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Sugar)[1] <- "Country" # Rename the first column
Sugar <- gather(Sugar, Year, "Sugar", 2:ncol(Sugar)) # Reshape dataframe
Sugar$Year <- factor(Sugar$Year, ordered = T) # Ordered column Year
Sugar[Sugar == 0.0] <- NA # Repalce all zeros with NA
Sugar<- Sugar %>% na.omit() # Omit na

#Merge data & Sugar
data_Sugar <- merge(data, Sugar, by = c("Country", "Year"))
data <- data_Sugar

# Read Food related .csv file, Reshape 
Food <- read.csv("indicator food_consumption.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Food)[1] <- "Country" # Rename the first column
Food <- gather(Food, "Year", "Food", 2:ncol(Food)) # Reshape dataframe
Food$Year <- factor(Food$Year, ordered = T) # Ordered column Year


# Merge data & Food
data_Food <- merge(data,Food, by = c("Country", "Year"))
data <- data_Food

# Read Income related .csv file, Reshape
Income <- read.csv("indicator gapminder gdp_per_capita_ppp.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Income)[1] <- "Country" # Rename the first column
Income <- gather(Income, "Year", "Income", 2:ncol(Income)) # Reshape dataframe
Income$Year <- factor(Income$Year, ordered = T) # Ordered column Year

# Merge data & Income
data_Income <- merge(data, Income, by = c("Country", "Year"))
data <- data_Income
data <- data %>% na.omit() # data without NA values

# Read "indicator life_expectancy_at_birth.csv"
Life<-read.csv("indicator life_expectancy_at_birth.csv", header = TRUE, check.names = FALSE) # Read .csv file
names(Life)[1] <- "Country" # Rename the first column
Life <- gather(Life, "Year", "Life_expectancy", 2:ncol(Life)) # Reshape dataframe
Life$Year <- factor(Life$Year, ordered = T) # Ordered column Year

# Merge data & Life dataframe
data_Life <- merge(data, Life, by = c("Country","Year"))
data <- data_Life

data <- data %>% na.omit() # Omit the NA values in data


countries_eu <- c("Albania", "Andorra", "Armenia", "Austria", "Azerbaijan", "Belarus", "Belgium", "Bosnia and Herzegovina", "Bulgaria", "Croatia", "Cyprus", "Czech Rep.", "Denmark", "Estonia", "Finland", "France", "Georgia","Germany", "Greece", "Hungary", "Iceland", "Ireland", "Italy", "Kazakhstan", "Latvia"," Lithuania", 
 "Luxembourg", "Macedonia, FYR", "Malta", "Moldova", "Montenegro", "Netherlands", "Norway", "Poland", "Portugal", "Romania", "Russia", "Serbia", "Slovak Republic", "Slovenia", "Spain", "Sweden", "Switzerland", "Turkey", "Ukraine", "United Kingdom") # EU countries 

nordic_c <- c("Denmark", "Finland", "Iceland", "Norway", "Sweden") # Northern countries 
south_c <- c("Cyprus","Greece", "Italy", "Malta", "Portugal", "Spain", "Turkey" ) # Southern countries
east_c <- c("Albania", "Azerbaijan", "Armenia", "Belarus", "Azerbaijan", "Bosnia and Herzegovina", "Bulgaria", "Croatia", "Czech Rep.", "Estonia", "Georgia", "Hungary",  "Kazakhstan", "Latvia"," Lithuania",  "Macedonia, FYR",  "Moldova", "Montenegro", "Poland", "Romania", "Russia", "Serbia", "Slovak Republic", "Slovenia", "Ukraine") # Eastern countries
western_c <- c("Andorra",  "Austria",  "Belgium", "France", "Germany", "Ireland", "Luxembourg", "Netherlands", "Switzerland", "United Kingdom") # Western countries


data_eu <- filter(data, Country %in% countries_eu) # Create the data set data_eu, which is subset of data. Contains only european countries
data_eu$Country <- as.factor(as.character(data_eu$Country)) # Convert variable Country to factor
data_eu$Year <- as.factor(as.character(data_eu$Year)) # Convert variable Year to factor
data_eu$Gender <- as.factor(as.character(data_eu$Gender)) # Convert variable Gender to factor

data_eu$Region <- NA # Set all values in Region to "NA"

data_eu$Region[data_eu$Country %in% nordic_c] <- "Northern Europe" # If Country located in Northern Europe set Region to "Northern Europe"
data_eu$Region[data_eu$Country %in% south_c] <- "Southern Europe" #  If Country located in Southern Europe set Region to "Southern Europe"
data_eu$Region[data_eu$Country %in% east_c] <- "Eastern Europe" # If Country located in Eastern Europe set Region to "Eastern Europe"
data_eu$Region[data_eu$Country %in% western_c] <- "Western Europe" # If Country located in Western Europe set Region to "Western Europe"

data_eu$Region <- as.factor(as.character(data_eu$Region)) # Convert variable Region to factor
     

# Create three periods in which the data_eu will be examined. Investigate the data_eu in one period many details can be discovered.

period_1 = c("1980", "1981", "1982", "1983", "1984", "1985", "1986", "1987", "1988", "1989")
period_2 = c("1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999")
period_3 = c("2000", "2001", "2002", "2003", "2004")

#Create variable Period
data_eu$Period[data_eu$Year %in% period_1] <- "1980 - 1989" # If Year in period_1 set Period to "1980 - 1989"
data_eu$Period[data_eu$Year %in% period_2] <- "1990 - 1999" # If Year in period_2 set Period to "1990 - 1999"
data_eu$Period[data_eu$Year %in% period_3] <- "2000 - 2004" # If Year in period_3 set Period to "2000 - 2004"

data_eu$Period <- as.factor(as.character(data_eu$Period)) #Covert Period to factor

```



```{r echo = FALSE}
write.csv(data_eu, file = "data_eu.csv", row.names=FALSE)
```

